# Data importation {#data-import}

This is a short exercise to show you typical ways of importing data into R,
using both **base R** and **tidyverse** functions.

## Setting up your data folder

First, let's create a folder named `data` in your working directory to store
all the data files, which you will download from Canvas (or from the links
below, generated for this book).

```{r}
#| label: setup-data-folder
#| eval: false
#| warning: false

dir.create("data")
```

```{r}
#| label: create-data
#| include: false
#| eval: true

# This chunk runs when the BOOK is rendered (not by students). It generates
# example data files and writes them into data/, which is listed under
# `resources:` in _quarto.yml so Quarto copies it into _book/data/ on
# render. That makes the files downloadable from the published GitHub
# Pages site at a relative link like `data/data.csv`.

dir.create("data", showWarnings = FALSE)

df <- data.frame(
  id     = 1:6,
  name   = c("Alice", "Bob", "Carol", "David", "Ellen", "Frank"),
  age    = c(25, 30, 22, 41, 36, 29),
  bmi    = c(22.1, 27.4, 19.8, 31.2, 24.6, 26.0),
  smoker = c("no", "yes", "no", "yes", "no", "no")
)

writeLines(as.character(df$age), "data/data.txt")
write.table(df, "data/data.tsv", sep = "\t", row.names = FALSE, quote = FALSE)
write.csv(df, "data/data.csv", row.names = FALSE)
writexl::write_xlsx(df, "data/data.xlsx")
```

<!-- Download these four files and move them into your local `data` folder:
[`data.txt`](../data/data.txt), [`data.tsv`](../data/data.tsv),
[`data.csv`](../data/data.csv), and [`data.xlsx`](../data/data.xlsx). -->
Download these four files and move them into your local `data` folder:
<a href="../data/data.txt" download>`data.txt`</a>,
<a href="../data/data.tsv" download>`data.tsv`</a>,
<a href="../data/data.csv" download>`data.csv`</a>, and
<a href="../data/data.xlsx" download>`data.xlsx`</a>.

Check the **Files** pane in RStudio. You should see a new `data` folder
appear inside your working directory, containing the four files. If they
don't show up, click the refresh icon at the top of the Files pane.

::: {.common-mistake}
If the `data` folder or files don't show up where you expect, double-check
you're looking inside your **project** folder, not some other location on
your computer. This is exactly the kind of confusion that RStudio Projects
and a consistent working directory are meant to prevent.
:::

## Base R import functions

Base R ships with several functions for reading plain-text data, with no
extra packages required.

- **`readLines()`** reads a text file line by line into a character vector.
  It doesn't know anything about columns or delimiters — it's the most
  "raw" way to bring text into R, useful for logs, unstructured text, or
  files you want to parse yourself.

  ```{r}
  #| eval: false
  lines <- readLines("data/data.txt")
  lines
  ```

- **`read.table()`** is the general-purpose workhorse for delimited files.
  You tell it the separator (`sep = "\t"` for tab-delimited, `sep = ","`
  for comma-delimited, etc.), and it returns a `data.frame`. Its defaults
  are geared toward "traditional" R usage: strings are sometimes converted
  to factors (depending on R version/options), and column names may be
  mangled to valid R names unless you tell it otherwise.

  ```{r}
  #| eval: false
  df_tsv <- read.table("data/data.tsv", sep = "\t", header = TRUE)
  str(df_tsv)
  ```

- **`read.csv()`** is simply `read.table()` with sensible defaults baked in
  for comma-separated files (`sep = ","`, `header = TRUE`). It's the
  function most people reach for first when reading a `.csv` in base R.

  ```{r}
  #| eval: false
  df_csv <- read.csv("data/data.csv")
  str(df_csv)
  ```

## Tidyverse import functions

The tidyverse (via the **readr** and **readxl** packages) offers analogous
functions with a more consistent, predictable, and faster implementation.

- **`read_csv()`** (from **readr**) reads comma-separated files into a
  `tibble` rather than a base `data.frame`. Tibbles print more compactly,
  never auto-convert strings to factors, guess column types explicitly
  (printing a helpful message about what it guessed), and preserve column
  names exactly as they appear in the file (including spaces, if any).

  ```{r}
  #| eval: false
  library(readr)
  df_csv_tv <- read_csv("data/data.csv")
  df_csv_tv
  ```

- **`readxl::read_excel()`** reads Excel spreadsheets (`.xls` or `.xlsx`)
  directly, without needing Excel installed or any conversion step. You can
  specify which `sheet` and which cell `range` to read, which is handy when
  a workbook has multiple tabs or extra header rows.

  ```{r}
  #| eval: false
  library(readxl)
  df_xlsx <- read_excel("data/data.xlsx")
  df_xlsx
  ```

### A note on the `::` notation

You may notice code like `readxl::read_excel()` instead of just
`read_excel()`, and `writexl::write_xlsx()` instead of just `write_xlsx()`.
The double-colon `::` is R's **namespace operator**: `package::function()`
tells R exactly which package a function should come from, without
requiring you to load that package with `library()` first.

This is useful for a few reasons:

- **Clarity.** Writing `readxl::read_excel()` makes it immediately obvious,
  just by reading the code, which package a function belongs to — handy
  when you're skimming a script months later or reading someone else's
  code.
- **Avoiding conflicts.** Sometimes two different packages define a
  function with the same name (for example, both **dplyr** and **stats**
  have a function called `filter()`). Using `::` lets you pick exactly the
  one you want instead of relying on whichever package was loaded most
  recently (and therefore takes priority).
- **One-off use.** If you only need a single function from a package, and
  don't want to load the whole package (which can be slow, or can clutter
  your environment with functions you don't need), `::` lets you call it
  directly, as we do above with `readxl::read_excel()` and
  `writexl::write_xlsx()`, without ever running
  `library(readxl)` or `library(writexl)`.

By contrast, for **readr**, we did run `library(readr)` before calling
`read_csv()` — this is a stylistic choice. Both approaches are valid: use
`library()` when you'll be calling many functions from a package throughout
your script, and use `::` when you only need a function once or twice, or
want to be explicit about where it's coming from.

### Base R vs. tidyverse: what's the difference?

Both families ultimately produce a rectangular table you can work with in
R, but they differ in a few practical ways: base R functions (`read.table`,
`read.csv`) return a `data.frame` and historically default to converting
character columns to factors and "cleaning" column names, whereas tidyverse
functions (`read_csv`, `read_excel`) return a `tibble`, keep strings as
strings by default, are generally faster on large files, and print explicit
messages about the column types they've guessed so you can catch parsing
mistakes early. Neither is "more correct."  Base R functions have no
dependencies beyond R itself (useful for portable scripts or package
development), while tidyverse functions are often preferred in interactive
data analysis for their speed and safer defaults.

::: {.exercise}
**Exercise 1.4**

Explore any one of the imported datasets with `head()`, `tail()`,
   `summary()`, `nrow()`, `ncol()`, `dim()`, and `str()`. Note down what
   each function tells you about the data.
:::
