# Data wrangling {#data-wrangling}

Real datasets are almost never as clean as `viz_data.csv` has been so far.
In this chapter you'll get comfortable with the messiness that shows up
in real public health data, learn how to check a dataset for problems
before you trust it, and build up a reproducible cleaning workflow using
some of the most common `dplyr` and `stringr` functions you'll reach for
again and again.

## Data setup

For this chapter we'll use a deliberately messy dataset,
[`clinic_raw.csv`](../data/clinic_raw.csv), styled after the kind of
export you might get from a clinic's Excel records. Download it and move
it into your `data` folder before continuing.

```{r}
#| label: create-data
#| include: false
#| eval: true

dir.create("data", showWarnings = FALSE)
set.seed(42)

# simulate a messy export: two junk header lines before the real header
# (mimicking a title row and a blank row, as Excel exports often have),
# inconsistent date formats, some values stored as "N/A" or "-" instead
# of true NA, extra whitespace and inconsistent case in text fields, and
# a mix of numeric fields stored as character due to stray symbols
n = 200
ids = sprintf("PT-%04d", 1:n)
sites = sample(c("  site a", "Site B", "SITE C ", "site_d"), n, replace = TRUE)
sex = sample(c("M", "F", "m", "f", NA), n, replace = TRUE, prob = c(0.35, 0.35, 0.1, 0.1, 0.1))
age = sample(c(18:90, NA), n, replace = TRUE)
visit_date = format(
  sample(seq(as.Date("2023-01-01"), as.Date("2023-12-31"), by = "day"), n, replace = TRUE),
  "%d/%m/%Y"
)

# blood pressure is recorded as "systolic/diastolic mmHg" in one messy
# text column, exactly the kind of single-field-holding-two-values
# problem you'll often hit with real clinic exports
systolic = sample(90:180, n, replace = TRUE)
diastolic = sample(55:110, n, replace = TRUE)
bp_raw = ifelse(
  runif(n) < 0.08, "N/A",
  paste0(systolic, "/", diastolic, " mmHg")
)

weight_kg = ifelse(runif(n) < 0.05, "-", as.character(round(rnorm(n, 70, 12), 1)))
notes = sample(c("stable", "Stable", "FOLLOW UP", "follow up", "  discharged", NA),
                 n, replace = TRUE)

clinic_df = data.frame(
  patient_id = ids,
  site = sites,
  sex = sex,
  age = age,
  visit_date = visit_date,
  blood_pressure = bp_raw,
  weight_kg = weight_kg,
  notes = notes
)

junk_lines = c("Clinic Export - Confidential", "")
out_path = "data/clinic_raw.csv"
writeLines(junk_lines, out_path)
write.table(clinic_df, out_path, sep = ",", row.names = FALSE, append = TRUE)
```

```{r}
#| label: import-data
#| message: false

# load the multiple core tidyverse packages all at once
library(tidyverse)

# skip = 2 drops the title line and blank line sitting above the real header
clinic_raw = read_csv("data/clinic_raw.csv", skip = 2)
head(clinic_raw)
```

Notice the `skip = 2` argument to `read_csv()`. The raw file has two junk
lines above the real header (a title line and a blank line), a common
situation when exporting from Excel, and `skip` tells R to ignore a
fixed number of lines at the top of the file before looking for column
names. If you only need specific columns or rows rather than skipping a
fixed block, `read_csv()`'s `col_select` argument and ordinary row
indexing (e.g. `clinic_raw[1:20, ]`) work the same way they did with
`viz_data.csv`.

::: {.pro-tip}
`1:20` is shorthand for `seq(1, 20, by = 1)`, a function that
generates a sequence of numbers from a starting value to an ending value
at a given increment. The colon operator (`:`) is just a quick way to
write this when your increment is `1` (or `-1` when counting downward),
but `seq()` is worth knowing too, since it lets you control the
increment directly, for example generating every second or every tenth
number in a range.
:::

```{r}
#| label: colon-vs-seq
1:20 # shorthand sequence from 1 to 20, increment of 1
seq(1, 20, by = 1) # explicit equivalent of 1:20
seq(1, 20, by = 5) # 1:20 can't do this, seq() lets you set the increment
```

## The messiness of real-world data

Two problems show up immediately if you inspect `clinic_raw`. First,
missing values aren't always stored as true `NA`. Here, `blood_pressure`
uses `"N/A"` and `weight_kg` uses `"-"` for missing entries, both of
which R reads in as ordinary text rather than recognising them as
missing. Second, columns that should be numeric, like `blood_pressure`
(which packs two readings and a unit into one field, e.g.
`"120/80 mmHg"`) and `weight_kg`, get read in as character columns
instead, a process called *coercion*, because a single non-numeric value
anywhere in the column forces the whole column to be stored as text.

```{r}
#| label: check-column-types
str(clinic_raw) # lists each column's name and type in one compact view
```

You'll fix both of these problems later in this chapter, once you've
seen the cleaning functions designed for exactly this job.

::: {.common-mistake}
It's easy to assume a column is numeric just because most of its values
look like numbers. Always check with `str()` or `class()` after
importing, since a single stray value like `"N/A"` or a trailing unit
like `"mmHg"` is enough to silently turn an entire column into
character.
:::

## Checking data quality

Before cleaning anything, it's worth getting an honest picture of what's
actually in your dataset.

- **`is.na()`** checks, value by value, whether an entry is missing,
  and is most useful combined with `sum()` to count how many missing
  values a column has.
- **`range()`** returns the minimum and maximum of a numeric variable,
  handy for spotting impossible values (e.g., a negative age).
- **`summary()`** gives you a quick five-number-summary-plus-mean for
  numeric columns, and a count of each category for character or factor
  columns, all in one call.

```{r}
#| label: count-missing-age
sum(is.na(clinic_raw$age)) # TRUE counts as 1, so sum() tallies the NAs
```

```{r}
#| label: check-age-range
range(clinic_raw$age, na.rm = TRUE) # na.rm drops NAs before computing min/max
```

```{r}
#| label: summarise-clinic-data
summary(clinic_raw)
```

Numbers alone don't always tell the full story, so it's worth pairing
these checks with a quick plot. A histogram of `age`, for instance, can
reveal implausible spikes or gaps that summary statistics alone might
not make obvious.

```{r}
#| label: plot-age-distribution
#| message: false

library(ggplot2)
ggplot(clinic_raw, aes(x = age)) +
  geom_histogram(binwidth = 5, na.rm = TRUE)
```

## Reproducible data cleaning

The rest of this chapter walks through a toolkit of functions for
turning `clinic_raw` into something trustworthy, one small, well-named
step at a time, so that anyone (including future you) can re-run your
script and get the exact same cleaned dataset back out.

### Renaming, creating, and selecting columns

- **`rename()`** changes a column's name without touching its values.
- **`mutate()`** creates a new column, or overwrites an existing one,
  based on some calculation.
- **`select()`** keeps (or drops, with a `-`) specific columns.
- **`filter()`** keeps only the rows that match a condition.

```{r}
#| label: rename-columns
clinic_clean =
    clinic_raw %>%
    rename(id = patient_id) # patient_id -> id, values untouched
```

```{r}
#| label: select-columns
clinic_clean =
    clinic_clean %>%
    select(id, site, sex, age, visit_date, blood_pressure, weight_kg, notes)
```

```{r}
#| label: filter-adults
clinic_clean =
    clinic_clean %>%
    filter(age >= 18 | is.na(age)) # keep adults, but don't drop missing ages
```

### Sorting rows

- **`arrange()`** sorts rows by one or more columns, smallest to
  largest by default.
- Wrapping a column in **`desc()`** inside `arrange()` sorts it
  largest to smallest instead.

```{r}
#| label: arrange-by-age
clinic_clean %>% arrange(age) %>% head() # youngest patients first
```

```{r}
#| label: arrange-by-age-desc
clinic_clean %>% arrange(desc(age)) %>% head() # oldest patients first
```

### Grouped summaries

- **`group_by()`** splits your data into groups based on one or more
  columns, without changing how it looks yet.
- **`summarise()`** then collapses each group down to a single summary
  row, using whatever calculation you specify.

```{r}
#| label: mean-age-by-site
clinic_clean %>%
    group_by(site) %>%
    summarise(
          mean_age = mean(age, na.rm = TRUE)
        , n = n() # count number of rows in each group
    )
```

### Cleaning up text

Notice that `site` has inconsistent spacing, capitalisation, and even an
underscore in one category, and `notes` has similar inconsistencies.
These are exactly the kind of problems the next set of functions fix.

- **`trimws()`** removes leading and trailing whitespace from text.
- **`toupper()`** and **`tolower()`** convert text to all-uppercase or
  all-lowercase, which is a simple way to make categories consistent
  when only their capitalisation differs.
- **`substring()`** extracts a portion of a string based on start and
  end character positions.
- **`sub()`** replaces the *first* match of a pattern in a string with
  something else, while its cousin `gsub()` replaces *all* matches.

```{r}
#| label: clean-site-column
clinic_clean =
    clinic_clean %>%
    mutate(
          site = tolower(trimws(site)) # lowercase + strip whitespace
        , site = gsub("_", " ", site)  # replace underscore with space
    )
table(clinic_clean$site)
```

```{r}
#| label: substring-example
substring("PT-0001", 4, 7) # characters 4 through 7: "0001"
```

### Regular expressions

Sometimes the pattern you want to match or extract is more flexible than
a fixed piece of text, that's where regular expressions come in, a
mini-language for describing patterns in text.

- **`str_extract()`** (from the `stringr` package) pulls out the part of
  a string that matches a pattern.
- **`gsub()`** can also take a regular expression as its pattern, not
  just a fixed string, making it useful for replacing anything that
  *looks like* a pattern (e.g., "any digits") rather than one exact
  value.

Recall that `blood_pressure` packs two readings into one field, like
`"120/80 mmHg"`. We can pull out the systolic (before the `/`) and
diastolic (after the `/`) readings separately using two different
regular expressions.

```{r}
#| label: extract-systolic-diastolic
#| message: false

library(stringr)
clinic_clean =
    clinic_clean %>%
    mutate(
          bp_upper = as.numeric(str_extract(blood_pressure, "^[0-9]+")) # digits before the "/"
        , bp_lower = as.numeric(str_extract(blood_pressure, "(?<=/)[0-9]+")) # digits after the "/"
    )
head(clinic_clean[, c("blood_pressure", "bp_upper", "bp_lower")])
```

```{r}
#| label: remove-unit-with-gsub
gsub("[^0-9.]", "", "72.5 kg") # strip everything except digits and "."
```

- **`grep()`** returns the *positions* (or, with `value = TRUE`, the
  actual values) of the elements in a vector that match a pattern.
- **`grepl()`** returns a `TRUE`/`FALSE` for every element, indicating
  whether it matches the pattern, which is especially handy inside
  `filter()`.

```{r}
#| label: grep-followup-notes
grep("follow", clinic_clean$notes, ignore.case = TRUE) # ignore.case catches "FOLLOW UP" and "follow up"
```

```{r}
#| label: grepl-followup-notes
clinic_clean %>% filter(grepl("follow", notes, ignore.case = TRUE))
```

### Converting between data types

Recall from earlier that `blood_pressure` and `weight_kg` were coerced
into character columns. Now that you've extracted `bp_upper` and
`bp_lower` as numeric columns above, `weight_kg` still needs the same
treatment, along with turning `"N/A"` and `"-"` into genuine `NA` values
and fixing up `visit_date`.

- **`as.numeric()`** and **`as.integer()`** convert text (or other
  numeric-like values) into numeric or whole-number form.
- **`as.character()`** converts a value into text.
- **`as.logical()`** converts a value into `TRUE`/`FALSE`, recognising
  strings like `"TRUE"`, `"FALSE"`, `"T"`, and `"F"`.
- **`as.Date()`** converts text into an actual date object, using a
  `format` argument to tell R how the text is currently laid out.

```{r}
#| label: clean-weight-and-date
clinic_clean =
    clinic_clean %>%
    mutate(
          weight_kg = na_if(weight_kg, "-") # turn placeholder "-" into real NA
        , weight_kg = as.numeric(weight_kg) # now safe to convert to numeric
        , visit_date = as.Date(visit_date, format = "%d/%m/%Y") # dd/mm/yyyy text -> Date
    )
head(clinic_clean)
```

::: {.common-mistake}
`as.numeric()` on a column that still has stray text in it (like `"N/A"`
or a unit such as `"mmHg"`) won't error, it will just silently produce
`NA` for every value it can't parse, along with a warning message that's
easy to miss. Always check `sum(is.na(x))` before and after a conversion
like this to make sure you haven't accidentally lost more data than you
expected.
:::

### Conditional logic

- **`ifelse()`** checks a condition for every element of a vector and
  returns one value if it's `TRUE` and another if it's `FALSE`.
- **`case_when()`** extends this same idea to any number of conditions,
  checked in order, which is much easier to read than nesting several
  `ifelse()` calls inside each other.

```{r}
#| label: ifelse-example
clinic_clean =
    clinic_clean %>%
    mutate(
        sex = ifelse(
              toupper(sex) == "M", "Male"
            , ifelse(toupper(sex) == "F", "Female", NA) # nested ifelse for the third outcome
        )
    )
```

```{r}
#| label: case-when-bp-category
clinic_clean =
    clinic_clean %>%
    mutate(
        bp_category = case_when(
              is.na(bp_upper)   ~ NA_character_ # keep missing as missing
            , bp_upper < 120    ~ "Normal"
            , bp_upper < 140    ~ "Elevated"
            , TRUE              ~ "High" # catch-all for anything left
        )
    )
table(clinic_clean$bp_category, useNA = "always")
```

### Writing your own functions

Once you notice yourself repeating the same few lines of cleaning code
on more than one column, it's worth wrapping that logic into your own
function instead of copying and pasting it. A function you write
yourself works exactly like `mean()` or `filter()`, you just define what
it does once, then call it by name wherever you need it.

```{r}
#| label: define-clean-text-function
clean_text = function(x) {
  x = trimws(x)  # strip whitespace
  x = tolower(x) # standardise case
  x
}

clean_text("  FOLLOW UP")
```

```{r}
#| label: apply-clean-text-function
clinic_clean =
    clinic_clean %>%
    mutate(notes = clean_text(notes)) # reuse the function on the notes column
table(clinic_clean$notes, useNA = "always")
```

### Reshaping data

- **`pivot_longer()`** takes several columns and stacks them down into
  two columns, a "names" column and a "values" column, making wide data
  longer.
- **`pivot_wider()`** does the reverse, spreading one column's values
  out into multiple new columns, making long data wider.

`bp_upper` and `bp_lower` are a natural pair to reshape here: they're
both numeric blood pressure readings taken at the same visit, just two
components of one measurement, so stacking them into a single `value`
column (tagged by a `reading` column saying which is which) actually
makes sense, unlike forcing together two columns of different kinds of
data.

```{r}
#| label: pivot-longer-example
clinic_clean %>%
    select(id, bp_upper, bp_lower) %>%
    head(4) %>%
    tidyr::pivot_longer(
          cols = c(bp_upper, bp_lower) # the two columns to stack
        , names_to = "reading"         # new column naming which reading it was
        , values_to = "value"          # new column holding the actual number
    )
```

```{r}
#| label: pivot-wider-example
clinic_clean %>%
    group_by(site, sex) %>%
    summarise(mean_age = mean(age, na.rm = TRUE), .groups = "drop") %>%
    tidyr::pivot_wider(names_from = sex, values_from = mean_age)
```

### Combining datasets

- **`cbind()`** and **`rbind()`** glue data frames together side by side
  (by column) or stacked on top of each other (by row), assuming the
  pieces already line up correctly.
- **`merge()`** joins two data frames together by matching values in a
  shared column, base R's way of doing what a database would call a
  join.
- **`left_join()`**, **`right_join()`**, **`anti_join()`**, and
  **`semi_join()`** (from `dplyr`) offer the same idea as `merge()` but
  are usually easier to read: `left_join()` keeps every row from the
  first (left) table and attaches matching information from the second,
  `right_join()` does the same but keeps every row from the second
  table, `anti_join()` keeps only rows from the first table that have
  *no* match in the second, and `semi_join()` keeps only rows from the
  first table that *do* have a match, without actually adding any of the
  second table's columns.

```{r}
#| label: create-site-lookup
site_lookup = data.frame(
      site = c("site a", "site b", "site c", "site d")
    , region = c("North", "North", "South", "South")
)
```

```{r}
#| label: left-join-site-region
clinic_clean =
    clinic_clean %>%
    left_join(site_lookup, by = "site") # attach region, keep every clinic_clean row
head(clinic_clean)
```

### Pipes: `%>%` vs. `|>`

You've been using `%>%` throughout this chapter (from the `magrittr`
package, loaded automatically with `dplyr`) to chain several steps
together by passing the result of one line into the next. Base R now
has its own built-in pipe, `|>`, which works the same way and doesn't
need any package loaded at all, since it's part of R itself from
version 4.1 onward.

```{r}
#| label: compare-pipes
clinic_clean %>% filter(bp_category == "High") %>% nrow()
clinic_clean |> filter(bp_category == "High") |> nrow()
```

In practice, the two behave almost identically for the everyday chaining
you've done in this chapter, so you'll see both used interchangeably
across R code you come across. `%>%` has a couple of extra features
(like letting you refer to the piped value with a `.` in more flexible
positions) that `|>` doesn't fully replicate, but for straightforward
step-by-step pipelines like the ones here, neither is objectively
better, and which one you use often just comes down to personal
preference or what a particular course or workplace has standardised
on.

## Exporting cleaned data

Once your data is clean, you'll want to save it out so you (or a
collaborator) don't have to repeat the cleaning steps every time.

- **`writeLines()`** writes plain text, line by line, to a file, useful
  for simple text output but not really designed for tabular data.
- **`write.table()`** writes a data frame to a text file, with
  arguments controlling the separator, whether to include row names,
  and so on, base R's general-purpose export function.
- **`write.csv()`** is `write.table()` with sensible defaults baked in
  for comma-separated files, the base R equivalent of `read.csv()`.
- **`write_csv()`** (from `readr`) does the same job as `write.csv()`,
  but is faster on large datasets and never writes row names by
  default, avoiding a common annoyance with the base R version.

```{r}
#| label: export-cleaned-data
#| eval: false

write_csv(clinic_clean, "data/clinic_clean.csv")
```

::: {.exercise}
**Exercise 2.2**

In a new script file,

1. Import `clinic_raw.csv` yourself using `read_csv()`, remembering to
   skip the junk lines at the top of the file, and use `str()` and
   `summary()` to check which columns have missing values or the wrong
   data type.
2. Clean at least three problems you find: fix a column with the wrong
   type (using `as.numeric()`, `as.Date()`, or similar), standardise a
   text column's capitalisation or spacing, and turn a placeholder like
   `"N/A"` or `"-"` into a true `NA`.
3. Use `group_by()` and `summarise()` to calculate the mean `age` (or
   another numeric variable of your choice) by `site`.
4. Export your cleaned dataset to a new CSV file using `write_csv()`.
:::
