# Summaries and Statistical Tests {#summaries-statistical-tests}

You've already learned how to import data and make plots in R.
Let's now move further to data analysis. In this
exercise, you'll learn how to summarise a dataset numerically and run common
statistical tests using simple R functions.

## Data setup

We will reuse the same data we used for visualisation last week, [`viz_data.csv`](../data/viz_data.csv).

```{r}
#| label: import-data
#| message: false

viz_data = readr::read_csv("data/viz_data.csv")
head(viz_data)
```

## Common data summaries

Before jumping into formal tests, it's always worth describing your data
with some simple numbers. These are common measures of center and spread, the building blocks you'll rely on for almost every analysis you do.

- **`mean()`** gives you the arithmetic average of a numeric variable, a
  quick sense of where your data is "centred".
- **`median()`** gives you the middle value when the data is sorted,
  which is often more useful than the mean when your data has outliers
  (e.g., a few unusually high weight readings).
- **`min()`** and **`max()`** return the smallest and largest values,
  handy for spotting data entry errors as much as for describing the
  variable itself.
- **`quantile()`** splits your data into percentiles, so you can see, for
  example, the value below which 25% or 75% of observations fall.
- **`sd()`** gives you the standard deviation, a measure of how spread
  out your values are around the mean.

```{r}
#| label: mean-median-summary
mean(viz_data$weight)
median(viz_data$weight)
```

```{r}
#| label: range-quantile-sd
min(viz_data$bmi)
max(viz_data$bmi)
quantile(viz_data$bmi)
sd(viz_data$bmi)
```

- **`cov()`** tells you whether two numeric variables tend to move
  together (positive), move in opposite directions (negative), or show no
  clear linear relationship (close to zero). Its exact value is hard to
  interpret on its own because it depends on the units of your variables.
- **`cor()`** standardises this idea into a value between -1 and 1, making
  it much easier to interpret and compare across variable pairs.

```{r}
#| label: covariance-height-weight
cov(viz_data$height, viz_data$weight)
```

```{r}
#| label: correlation-height-weight
cor(viz_data$height, viz_data$weight)
```

## Common statistical tests

Now that you can describe your data, let's move on to running some tests
in R. You've already covered the logic behind these tests in your stats
class, so here we'll focus purely on how to actually call them in code
and read what they give you back.

### Association of two categorical variables

- **`chisq.test()`** tests whether two categorical variables are
  associated with one another, by comparing observed counts in a
  contingency table to what you'd expect if the variables were
  independent.
- **`fisher.test()`** answers the same underlying question, but uses an
  exact calculation rather than an approximation, which makes it more
  reliable when your contingency table has small cell counts.

```{r}
#| label: chi-squared-test
group_outcome_table <- table(viz_data$group, viz_data$outcome)
chisq.test(group_outcome_table)
```

```{r}
#| label: fisher-exact-test
fisher.test(group_outcome_table)
```

### Difference between two groups

- **`t.test()`** compares the means of a numeric variable between two
  groups, and assumes that variable is roughly normally distributed
  within each group.
- **`wilcox.test()`** (which performs the Mann-Whitney test when comparing
  two independent groups) compares the distributions of a numeric
  variable between two groups without assuming normality, making it a
  useful backup when that assumption feels shaky.

```{r}
#| label: t-test-response-dose
t.test(response ~ group, data = viz_data, subset = group %in% c("High dose", "Low dose"))
```

```{r}
#| label: mann-whitney-response-dose
wilcox.test(response ~ group, data = viz_data, subset = group %in% c("High dose", "Low dose"))
```

### ANOVA

- **`aov()`** extends the idea behind the t-test to three or more groups,
  testing whether the means of a numeric variable differ across the
  levels of a categorical variable.

```{r}
#| label: anova-response-site
response_site_aov = aov(response ~ site, data = viz_data)
summary(response_site_aov)
```

## Functions in R: base vs. additional packages

Every function you've used so far in this chapter, like `mean()`,
`t.test()`, and `aov()`, comes bundled with base R, meaning it's available
the moment you open RStudio with no extra installation needed. As you
progress in public health analysis, you'll also encounter functions from
additional packages (installed with `install.packages()` and loaded with
`library()`, just like you did with `stringr` last week) that
extend what base R can do.

## Accessing values from output objects

When you run a test like `t.test()` or `chisq.test()`, R doesn't just
print a result to your console. It actually creates a list-like object
containing everything about that test, and the printed summary is just
a friendly display of that object. Other functions, like `summary()`,
return a named vector rather than a list. This matters because in your
own code you'll often want to pull out a single number (like a p-value
or a specific summary statistic) to use later, rather than manually
reading it off the printed output, and the way you extract that number
depends on what kind of object you're pulling it from.

- **`str()`** lets you peek inside an output object to see what pieces of
  information it contains and how they're named, and whether it's a list
  or a vector.
- The `$` operator extracts one specific piece from a list-like object by
  name, once you know what it's called.
- `[` and `[[` let you extract by name or position from a named vector,
  which is what functions like `summary()` give you back instead of a
  list.

```{r}
#| label: inspect-test-object
response_ttest = t.test(response ~ group, data = viz_data, subset = group %in% c("High dose", "Low dose"))

str(response_ttest)

response_ttest$p.value
response_ttest$conf.int
```

```{r}
weight_summary = summary(viz_data$weight)

str(weight_summary)

weight_summary["Median"]
weight_summary[[3]]
```

::: {.common-mistake} 
It's tempting to guess the name of the value you want (e.g., typing $pvalue instead of $p.value), but R won't warn you if you get it wrong. It will just silently return NULL. It's also easy to try $ on something like the output of summary(), which will fail since that object is a named vector, not a list. Running str() first to check both the exact names and the type of object you're dealing with will save you this frustration. :::
:::

::: {.exercise}
**Exercise 2.1**

1. Compute the 80% interpercentile range for `bmi`. (Hint: `quantile`, output extraction, subtraction).
2. Check what the default alternative hypothesis is for `t.test`. Try performing a one-sided t-test.

:::
